Effects of bariatric surger and weight loss on resting-state functional connectivity

### 2020-12-02 intro_rsfmri.Rmd
introduction to manuscript saved and edited separately

---
title: "Effects of bariatric surger and weight loss on resting-state functional connectivity"
author:
- Hannah Heinrichs* [1]
- Frauke Beyer* [1][2]
- Kerstin Prehn 
- Agnes Floel 
- Arno Villringer [1][2][3]
- A. Veronica Witte [1][2][3]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book: 
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options: 
  chunk_output_type: console
---
```{r "To do", include=FALSE}

@@

#set wd (change depending on user)
setwd("/data/pt_02161/Analysis/Project2_resting_state/seed-based/Second_level /code_and_manuscript/")

@@

```
\newpage
# Abstract

\newpage
# Introduction
<!--Suggested Order of Introduction
1. Obesity is a major health concern. [... some more details about metabolic and mortality risk, especially related to morbid obesity ...] More and more recognized, midlie obesity also impacts later cognitive decline and dementia risk.
Obesity has been associated with differences in brain structure and function.
Regarding structural alterations, longitudinal studies have indicated that higher body mass index is related to increased cortial thinning in midlife (path through life study), potentially mediated by inflammation or glucose dysregulation... Similarly, white matter microstructure and may be related to adverse metabolic effects (lampe)
Differences in brain function in obesity have been reported in the reward network during task-based fMRI, highlighting the role of reward processing in obesity development and maintainance.
Resting state fMRI studies have also shown differences in connectivity in diverse brain networks, including reward and salience network, visual networks and the default mode network. In part, these differences have been related to metabolic and behavioral differences, yet there may be an important confounding due to head motion.
In this study, we aim to investigate the impact of bariatric surery on..
-->

<!-- obesity and interventions -->
Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.
_something about comorbidities_
Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. 
Luckily, dietary interventions such as caloric restriction and related metabolic improvements, and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
In some cases behavioral intervention fail to yield the desired success regarding weight loss. There are non-behavioral alternatives at hand.
<!-- bars most invasive (brain metabolism), last resort -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions, and co-morbidities, such as diabetes [@]. Roux-en-Y gastric bypass (RYGB) is the most common procedure as it not only restricts the food intake by reducing stomach volume but also affects the absorption of macronutrients and the hormonal profile [@]. <!-- ref for "common", ref for procedure -->
Other widely used procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. <!-- mehr zu Veränderung in Gasto-intestinaler Physiologie -->

<!-- Neural mechanisms/concurrent plasticity? -->
Precise mechanisms are yet to be elucidated. As a first step, a comprehensive understanding of the concurrent physiological changes underlying weight loss and recovery is needed. This will inform inferences about potential mechanisms and either support interventions proven to be effective or ideally allow imitating them will less invasive alternatives.

Recently, research focuses have slowly shifted to investigate brain metabolism structural and functional changes after surgery.
<!-- severe nutritional stress [@Crujeiras_2017] -->

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in the obese population, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
<!-- mechanism reward and appetite, overating -->
Besides malabsorption and restricted food ingestion, patients report post-surgical changes in food preferences and eating behavior. Those psychophysiological and behavioral effects may be related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which may influence weight regulation [@Mulla_2017, @Nudel_2019].
<!-- results in lean -->
The reward network entailing the ventromedial prefrontal cortex (vmPFC), the striatum (STR) but also the anterior insula (antIns) has been linked to food valuation processes in decision-making in lean individuals [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018; @Wiemerslage_2016].
<!-- results in obese longitudinal -->
Similarly, several studies have found differences in resting-state activity in reward regions depending on weight status.

Obese individuals exhibited increased connectivity between frontal and mesolimbic regions and weaker connectivity within regions of the mesolimbic system (encompassing limbic and striatal regions). These differences seemed to normalize, as trends reversed when looking at the resting state of obese individuals after compared to before the surgery [@Duan_2020].

On the other hand, BMI prior to the surgery was associated with changes in connectivity [@Duan_2020].

s concerning such processes following a surgical intervention are inconclusive.

@@

<!-- Default mode network --> 
Weight regulation and energy expenditure is dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. Obesity may be cause or effect of a decompensation of the psychophysiological system that ensures homeostasis.
The default mode network is commonly associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> It is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- difference lean obese --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2016]. Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
A longitudinal study that confirmed the stronger connectivity (of PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG). However, default mode connectivity in these regions seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable 1 year after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
This is consistent with other findings of reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC [@Li_2018] <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery.

The exact mechanisms of how bariatric surgery affects functional plasticity needs to be investigated. Among patients undergoing surgery, change in BMI was negatively correlated with the changes in connectivity of the default mode network, namely between the PCC/precuneus and the DLPFC [@Li_2018].

On the other hand, connectivity also served as a predictor of weight loss after surgery.

On the other hand, there are also indicators of reduced functional integrity, e.g. for the anterior cingulate cortex (ACC) at resting state compared to lean individuals [@Kullmann_2011].

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network (DMN).
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008].

@@

\newpage
# Methods

```{r Load data, include=FALSE}
source("get_subject_ID_group_tp_forSwE.R")
final = save_txt_for_swe(group = "all", tp = "all")

# sample size
#nrow(subj_df)
subj_idx <- match(unique(final$subj.ID), final$subj.ID) # indices of first row of each subject
subj_df <- final[subj_idx,] # df with unique subjects

# completed time points
freqtable <- as.data.frame(table(final$subj.ID))
tpfreq <- table(freqtable$Freq)

# overview distribution of data points
tp1 <- as.character(freqtable$Var1[freqtable$Freq == 1])
tp2 <- as.character(freqtable$Var1[freqtable$Freq == 2])
tp3 <- as.character(freqtable$Var1[freqtable$Freq == 3])
tp1_tab <- table(subset(final, subj.ID %in% tp1)$tp, subset(final, subj.ID %in% tp1)$condition)
tp2_tab <- table(subset(final, subj.ID %in% tp2)$tp, subset(final, subj.ID %in% tp2)$condition)[c(1,3),]
tp3_tab <- table(subset(final, subj.ID %in% tp3)$tp, subset(final, subj.ID %in% tp3)$condition)[1,]
total_datapoints <- c(sum(final$group==1),sum(final$group==2))
total_subjects <- plyr::count(final[subj_idx,"group"])$freq
tab_sample <- data.frame(rbind(tp1_tab, tp2_tab, tp3_tab, total_subjects, total_datapoints))
rm(tp1,tp2,tp3,tp1_tab, tp2_tab, tp3_tab, total_subjects, total_datapoints)
rownames(tab_sample) <- c("count: only 0","count: only 6","count: only 12","count: 0 and 6","count: 6 and 12","count: complete data","total number of subjects","total data points")
colnames(tab_sample) <- c("BARS","NBARS")

# distribution of condition and sex
subj_df$Sex <- plyr::mapvalues(subj_df$Sex, from = c("-1", "1"), to = c("f", "m"))
cs <- addmargins(table(subj_df$Sex,subj_df$condition))

dfw <- tidyr::pivot_wider(data = final,
                          id_cols = c("subj.ID","condition","Sex","Age_BL"),
                          names_from = "tp",
                          values_from = c("BMI","logmFD"))
```
```{r tabsample, echo=FALSE}
tab_sample <-
  knitr::kable(
    tab_sample,
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    caption = "Distribution of data points at months after intervention")
tab_sample
```

## Sample and study design

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/mzyxv](https://osf.io/mzyxv).
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. 
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study. <!-- Frage: Ausschluss bei head movement? ADI009_fu2 nicht enthalten (siehe Readme preprocessing) -->
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tabdescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tabsample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals (n = `r nrow(subj_df)`, `r nrow(filter(subj_df,Sex == "f"))` female; ages `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), that either underwent surgery (n = `r cs["Sum","KG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female). The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r test wd}
print(getwd())
```

```{r baseline characteristics, include=FALSE}
dfBL <- save_txt_for_swe(group = "all", tp = "BL")

# distributions per group
# BMI 
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()

tab_age <- psych::describeBy(data=dfBL, x=dfBL$Age_BL, group=dfBL$condition, mat=T)

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) + 
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). 
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tabdescr, echo=FALSE}

# table for descriptive statistics
descr <- psych::describeBy(dfw[,5:ncol(dfw)], group=dfw$condition,mat=T)
descr$predictor <- c(rep("BMI",6),rep("logmFD",6))
descr$tp <- car::recode(descr$vars, "c(1,4)='fu2';c(2,5)='fu';c(3,6)='bl'")
descr$group <- descr$group1
descr <- descr[,c("predictor","tp","group","n","mean","sd")]
descr <- descr[with(descr, order(predictor, tp)),]
descr[,c("mean","sd")] <- round(descr[,c("mean","sd")],2)
descr[c(2:6,8:12),"predictor"] <- ""
descr[c(2,4,6,8,10,12),"tp"] <- ""
rownames(descr) <- NULL

tab_descr <-
  knitr::kable(
    descr,
    row.names = FALSE,
    format = "latex",
    booktabs = T,
    caption = "BMI and mean FD of participants acquired in each condition and time point")

tab_descr

```
<!-- STUDY PROCEDURE -->
Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for STANDARDIZED?? breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

```{r fig_BMIdescr, fig.align='center', out.width="90%"}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
condition.labs <- c("BARS","NBARS") 
names(condition.labs) <- c("IG","KG")
# BMI over time each group
fig_BMIdescr <-
  ggplot(final, aes(
    x = tp,
    y = BMI,
    group = subj.ID,
    color = condition
  )) +
  geom_line() + geom_point() +
  stat_summary(
    aes(group = condition),
    fun = mean,
    geom  = "line",
    size = 2,
    color = c(rep("#00305E", 3), rep("#C00045", 3))
  ) +
  facet_grid(. ~ condition, labeller = labeller(condition=condition.labs)) + xlab("time points")
fig_BMIdescr +
  scale_x_discrete(labels = c("bl" = "0", "fu" = "6", "fu2" = "12")) +
  scale_color_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")
  ) +
  xlab("Time after intervention [months]") + ylab ("BMI [kg/m²]") +
  theme_bw() +  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "none"
  )
```

## fMRI acquisition

MRI images were retrieved with a 12-channel head coil of a 3 Tesla Siemens Trio system.
<!-- MRI --> T1-weighted anatomical images were acquired as described in @Prehn_2020 (with MPRAGE, TR = 1900 ms, TE = 2.52 ms, a = 9, voxel size = 1 x 1 x 1 mm$^3$, 192 sagital slices).
<!-- rsfMRI --> Resting state gradient echo were acquired in 34 slices at an 90$^\circ$ flip angel with a repetition time (TR) of 2.3 s and echo time (TE) of 30 ms; image matrix = 64 x 64 voxel with in-plane resolution of 3 mm x 3 mm and a slice thickness of 4 mm, 150 volumes<!-- dimension = [60;72;60] ??? -->. The total acquisition time was 5:45 minutes. Field map or phase-encode reversed images were not acquired.
Participants were instructed to close their eyes but to remain awake during scanning.

## Pre-proceessing

Data preprocessing was conducted with FSL and Freesurfer using a nipype (version 14.0.2) workflow implemented in python 2.7. 
The preprocessing workflow included anatomical, functional and diffusion-weighted imaging preprocessing and wrapped modules from FSL version 6.0.1<!-- ?! -->, Freesurfer version 6.0.0p1, ANTS <!-- version?! and so on-->. Details have been published on github [https//:.???.com]()
<!-- anatomical preprocessing write something about anatomical preprocessing (Freesurfer etc.) here --> 
The T1-weighted image was segmented to remove non-brain tissue and together with the functional images normalized to a default brain template from the Monteal Neurological Institute (MNI) space. Functional images were smoothed with an $0$ mm FWHM Gaussian kernel.

<!-- minimal functional preprocessing --> Preprocessing pipeline included removal of first four volumes and a transformation of the blood oxygen level dependent (BOLD) time series into the subject's anatomical space. This transform combines parameters from motion correction (FSL's MCFLIRT), fieldmap distortion correction (based on fieldmap and FSL's FUGUE) and coregistration to the subject's anatomical template space (using Freesurfer's bbregister and the transform from individual to longitudinal template space in Freesurfer's longitudinal processing stream).

<!-- Aroma + CC/ + GSR --> Volumes were further preprocessed with ICA-AROMA and compCor (CC) and, optionally, with a subsequent global signal regression (GSR) [@Ciric_2017; @Parkes_2018] to reduce the effect of physiological noise and head motion on the connectivity estimates.
As global signal regression is very efficient in removing the correlation of head motion and connectivity dependence, but is known to introduces spatial dependency and spurious correlations, we performed GSR after ICA-AROMA in a separate sensitivity analysis.

### Seed region of interest (ROI)
<!-- Regions of interest (definition and transformation into freesurfer long space) --> Seed regions were ... mm spheres in the precuneus (MNI: x,y,z = x,x,x) associated with the default mode network, and the Nucleus Accumbens (MNI: x,y,z = x,x,x) by averaging the time series across voxels.

### Connectivity maps
<!-- coregistration + normalization --> Voxel-wise seed-based connectivity maps in the individual subject's space (which was coregistered to the Freesurfer longitudinally preprocessed timepoint (more details)) were calculated using Pearson's correlation between average seed time series from pre-defined ROIs within the reward network (Nucleus accumbens, NAcc) and the DMN (precuneus, PCC). Subsequently, r-to-z transform was applied to the correlation to obtain normally distributed functional connectivity (using nilearn, nipype and python). 
Then, we will transform the connectivity maps into MNI space with a non-linear warp derived from anatomical preprocessing of the longitudinally preprocessed timepoints (using ANTS).

### Quality Assessment

<!-- EXCLUSION CRITERIA --> Participants in which head motion artifacts affected the T1-weighted image preprocessing (e.g. caused Freesurfer to fail or perform very poorly, i.e. rated "Failed" according to @Klapwijk_2019) were excluded because a subject's motion is highly similar across scans and we thus suspect the fMRI data to also suffer from motion artifacts in these subjects. 
As the expected average head motion is high [@Hodgson_2016], we did not exclude any other participants based on head motion estimates because it is difficult to determine a critical threshold in this sample. 
<!-- MOTION CORRECTION --> We therefore apply best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitor the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. As quality checks statistical analysis were performed on both with and without global signal regression, considering the controversial discussion about GSR introducing negative correlations and spatial bias into connectivity data [@Ciric_2017; @Murphy_2017].
Additionally, mean framewise displacement (FD) was extracted from 6 translational/rotational motion parameters of the whole time series according to @Power_2012, log-transformed and examined within and separate from the main models.

### Outcomes

The main outcome measure was the aggregated functional connectivity (FC) values for the two ROI NAcc and PCC.

## Analysis

Statistical analysis were performed in MATLAB using the sandwich estimator (SwE) toolbox 2.2.1 [@Guillaume_2014] as implemented in SPM12 and R version 3.6.1 [@team2013r].
During analyses an implicit mask and an explicit mask were used. Therefore, the MNI ICBM 2009a Nonlinear Symmetric brain mask [http://www.bic.mni.mcgill.ca/ServicesAtlases/ICBM152NLin2009](http://www.bic.mni.mcgill.ca/ServicesAtlases/ICBM152NLin2009) were re-sampled to 3 x 3 x 3 mm$^3$ and binarized at threshold $0.5$.

<!-- MODEL SETUP --> Two theoretically derived models were set up. The first model was aimed at testing the interaction of time and group: $FC = group + time + group*time$. The second model on the influence of between-subject differences in average BMI and longitudinal within-subject change in BMI was specified as follows: $FC = between-subject BMI + within-subject BMI$. 
These were examined each for correlation maps of two different seed regions (NAcc / PCC), and for the purpose of comparison for correlation maps of two different preprocessing pipelines (only AROMA+CC / with GSR)), with two different covariate sets (with/ without mean FD), resulting in $16$ a priori planned analyses.
Initially, analyses were run with a brain mask but after post hoc considerations detailed below, models were re-run with a grey matter mask which reduced the number of tests (voxels of interest) from $52378$ to $29451$. As a result a total of $32$ different analyses were performed.
Computation with the grey matter mask were exploratory and deviate from the preregistration.

<!-- Description marginal model --> The marginal model implemented via the SwE toolbox implicitly accounts for random effects without the need to specify them through the error term. It therefore accommodates any repeated measurement covariance structures. We used a modified SwE assuming different covariance structures for the intervention and the control group because of their unbalanced sample size, so heteroscedasticity would be considered.

<!-- Design matrix, vielleicht beide auch als Abbildung kurz zeigen? --> For the first model, time was introduced as factor because did not assume strict linearity for the increase in connectivity and sought to increase power to detect specific trends. The resulting flexible factorial design <!-- check: difference from full factorial or other? --> contained one regressor for each time point per group. For the second and third model, we calculated the centered mean BMI and log mean FD per subject and the intra-subject-centered BMI as proposed by @Guillaume_2014.
With a model containing average BMI and longitudinal change per subject, we investigated potential effects .............. on the complete sample and subsequently in a exploratory fashion for the intervention group only as the effect of longitudinal change in BMI was expected to be driven by this subsample.
<!-- Nuisance covariates --> Age at baseline, sex, and log mean framewise displacement (FD) were entered into the first and second model as nuisance variables. Age and sex were between-subject covariates while FD is time-varying. Mean FD was log-transformed and age was demeaned by subtracting the overall mean across subjects and time points. As change in FD and change in BMI share variance [@Beyer_2019], the inclusion of log mean FD may be disputed and done optionally.
<!-- Post hoc analyses  --> For that reason, we performed each analysis separately one with mean FD and without. Differences in results will be reported. For a better understanding of the unique contribution of average and longitudinal change FD measures, we employed a third post hoc model to disentangle the effects: $FC = between-subject FD + within-subject FD$. These post-hoc analyses deviate from the preregistration and should be considered to be exploratory.

<!-- Inference --> All parametric inferences on activation were drawn based on False Discovery Rate (FDR) adjusted p-values $\leq 0.05$. To ensure robustness of our results, we computed estimates with non-parametric test using Wild bootstrap with an unrestricted SwE on all contrasts of interest for voxelwise inference, p-values were FDR-corrected for $\alpha = 5%$.
In contrast to the parametric estimation, Wild Bootstrap avoids any parametric distributional assumptions (e.g. random field theory), and is unlike permutation based framework appropriate <!-- ??? clarify why--> for longitudinal data.

# Results

## Manipulation check 
<!-- check network presence -->

<!-- singletp (t-tests), nur diesen Check für BL reporten (gib dann oben noch an, dass du dafür SPM verwendet hast) --> We ran separate one-sample t-test for network activation at baseline controlling for log mean FD, age, and sex for each time point to check whether the assumed reward and default mode network was at all identifiable.

## Interaction effect of intervention and time

>>>>

---
title: "Effects of bariatric surger and weight loss on resting-state functional connectivity"
author:
- xxx [1]
- xxx [1][2]
- xxx [1][2][3]
- xxx [1][2]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book: 
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options: 
  chunk_output_type: console
---
```{r "To do", include=FALSE}

@@

@@

```
# Abstract

\newpage
# Introduction

<!-- obesity and interventions -->
Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.
Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. 
Luckily, dietary interventions such as caloric restriction and related metabolic improvements, and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
In some cases behavioral intervention fail to yield the desired success regarding weight loss. There are non-behavioral alternatives at hand.
<!-- bars most invasive (brain metabolism), last resort -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions, and co-morbidities, such as diabetes [@]. Roux-en-Y gastric bypass (RYGB) is the most common procedure as it not only restricts the food intake by reducing stomach volume but also affects the absorption of macronutrients and the hormonal profile [@]. <!-- ref for "common", ref for procedure -->
Other widely used procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. <!-- mehr zu Veränderung in Gasto-intestinaler Physiologie -->
<!-- Neural mechanisms/concurrent plasticity? -->
Precise mechanisms as to how bariatric surgery leads to recovery are yet to be elucidated. As a first step, a comprehensive understanding of the concurrent physiological changes underlying weight loss and recovery is needed. This will inform inferences about potential mechanisms and either support interventions proven to be effective or ideally allow imitating them with less invasive alternatives.

<!-- severe nutritional stress [@Crujeiras_2017] -->

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in the obese population. Thus, one recently emerging research focus is how brain metabolism and functional connectivity changes in response to surgery, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
The reward network comprising the ventromedial prefrontal cortex (vmPFC), the striatum (STR) but also the anterior insula (antIns) has been linked to food valuation processes in decision-making in lean individuals [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018; @Wiemerslage_2016]. These regions also seem to be altered after bariatric surgery [@Lips_2014; @Hogenkamp_2016; @Garcia_2012]. <!-- Zitationen checken -->
<!-- Normalization of connectivity --> Obese individuals exhibited increased connectivity between frontal and mesolimbic regions and weaker connectivity within regions of the mesolimbic system (encompassing limbic and striatal regions). But these differences seemed to normalize, as trends reversed when looking at the resting state connectivity after compared to before the surgery [@Duan_2020]. Resting state activitiy increased in subcortical regions, such as the insula and striatum<!-- specifically putamen --> [@Wiemerslage_2016]. 
<!-- mechanism reward and appetite, overeating --> Interestingly, patients report changes in food preferences and eating behavior after the surgery that may influence weight loss beyond malabsorption of nutrients and restricted food ingestion. Those psychophysiological and behavioral effects are presumably related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which may influence weight regulation [@Mulla_2017; @Nudel_2019].
<!-- results in lean -->

<!-- results in obese longitudinal -->
<!-- Similarly, several studies have found differences in resting-state activity in reward regions depending on weight status.  -->

<!-- On the other hand, BMI prior to the surgery was associated with changes in connectivity [@Duan_2020]. -->

<!-- s concerning such processes following a surgical intervention are inconclusive. -->

@@

<!-- Default mode network --> 
Because multiple resting state networks are relevant to obesity and associated with abnormal neurocircuitry interactions [@Ding_2020], extensive research has been devoted to investigating the the relationship between the control network and the reward network. However, the dynamics within the default network may also be of importance because it is associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> 
Weight regulation and energy expenditure, on the other hand, are highly dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. 
<!-- Obesity = decompensation of the psychophysiological system that ensures homeostasis. -->

The default mdoe network is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- study results --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2017].
Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
Comparing spontaneous activitiy before and after surgery shows a decrease in areas including superior frontal regions<!-- gehört zu SFC -->, and a decrease superior parietal lobule when in sated state [@Wiemerslage_2016].
Concerning spontaneuous functional connectivity in these regions, patient having undergone bariatric surgery resemble normal-weight individuals rather than obese individual in having a weaker connectivity within the default mode network [@Frank_2013].
A longitudinal study investigating confirmed that while there was stronger connectivity in the default model network <!-- (PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG) --> in obese patients before the surgery, it seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
Consistent with these findings, other studies reported a reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery [@Li_2018].

<!-- The exact mechanisms of how bariatric surgery affects functional plasticity needs to be investigated and especially the role of weight and weight change. -->
... 
It has been stated that among patients undergoing surgery, change in BMI was<!-- negatively --> correlated with the changes in connectivity of the default mode network, namely between the PCC/precuneus and the DLPFC [@Li_2018].
On the other hand, there are also indicators of reduced functional integrity, e.g. for the anterior cingulate cortex (ACC) at resting state compared to lean individuals [@Kullmann_2011]. ...

<!-- On the other hand, connectivity also served as a predictor of weight loss after surgery. -->

<!--  -->

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network (DMN).
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we will consider that different courses in increase and decrease of functional connectivity that are not necessarily linear are possible over the course of one year, depending on the phase of weight management [@Olivo_2017].

@@
